Reconstructing the last known movements of one of Nature's giants

### 2018-01-03 manuscript/revision/blue-whale-ms-v2.tex
Edits from CT as far as Fig 1

\newpage
\section{Main text}
Migratory species pose a particular challenge for conservation practitioners, because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
Such species may also be particularly affected by climate change, as changes can impact various parts of their life cycle\cite{robinson2009travelling}. 
Dealing with these issues is difficult due to the relative scarcity of information on individual-level movements over multiple years, for both historical and present-day populations, especially for wide-ranging marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.

The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
Despite blue whale hunting being banned for over 50 years, blue whales are still listed as Endangered by the IUCN Red List\cite{reilly2008balaenoptera}.
Conservation efforts are hampered by the scarcity of detailed data on the distribution, behaviour and ecology of blue whales, particularly for juveniles. 
Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}. 
These methods are adding vital behavioural data to our understanding of whale movements, however, they are often expensive, limited to short-term studies on few individuals\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}. 
Additionally, they do not provide information on historical movement patterns that could be important for understanding the drivers of whale declines.
An alternative technique for investigating blue whale movements is to use stable isotopes\cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioural information that are often under-utilised \cite{lister2011natural}.
We can use the isotopic compositions of whale tissues to reconstruct their movement patterns, by comparing them to spatially referenced models of isotopic variation across the species' range\cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}. 
These methods can be applied fairly inexpensively to large numbers of individuals, but as proof of concept, here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale `Hope'.

Baleen stable isotopes have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}.
Baleen is formed of plates of keratin and is used by mysticete whales to filter feed. 
It is ideal for stable isotope studies because keratin grows continuously through an individual's life (although it is worn away at the tips over time, so rarely records an individual's entire lifespan), and once laid down it is metabolically inert\cite{best1996stable}. 
Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around three weeks prior to that time (this time is required for the carbon and nitrogen to be assimilated and converted into tissues). %REF???
% add line about it being ok to use phytoplankton?

We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth. 
This resulted in XX discrete samples of baleen. % add number
We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin.
These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e. relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope\cite{west2006stable}.
Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; carbon isotopes provide information on the geographical locations of an individual, whereas nitrogen isotopes provide trophic and physiological information\cite{west2006stable}.

Given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$ (see Supplementary Methods), we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over the seven full years of the whale's life (early 1884 - spring 1891). 
The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods).  
In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig. \ref{fig1}; Figs. S1 and S2). 
In behavioural phase two (summer 1886 to spring 1891) $\delta^{15}$N values are relatively elevated and we see coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values. 
In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive $\delta^{13}$C values in the second half of 1891. 
The final three months of the record show a continued fall in $\delta^{13}$C values (Fig. \ref{fig1}; Figs. S1 and S2).

% figure 1
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis). Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes). There is strong annual periodicity and cross-correlation (Figs. S1 and S2) in both isotopes. The approximate relationship to years assuming a growth rate of 13.5 cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  \label{fig1}
\end{figure}

>>>>

\newpage
\section{Main text}
Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites\cite{runge2014conserving}. 
Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle\cite{robinson2009travelling}. 
Identifying threats to migratory species and developing effective conservation measures requires information on individual-level movements spanning multiple years, for both historical and present-day populations. Such data are scarce, especially for rare, wide-ranging and long-lived marine species\cite{ryan2013stable,hall2005stable,bailey2009behavioural}.

The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
Despite hunting for blue whales being banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List\cite{reilly2008balaenoptera}.
Blue whales in the northeast Atlantic appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations (REF).  
The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland\cite{pike2009note}.
Their small population size and oceanic habit makes blue whale movements extremely hard to study. 
Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands\cite{baines2014upwellings}.
Northward migration in the spring may occur along mid-Atlantic corridors with peak sightings in the Azores around April-May. Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas\cite{pike2009note}.
Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals\cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
While direct tagging methods are adding vital behavioural data to our understanding of whale movements\cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies\cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour\cite{walker2012review}.
Currently the longest period of continuous monitoring via satellite tagging of an individual blue whale in the northeast Atlantic is X days, while photographic identification has matched a single blue whale with sightings in Iceland and Mauritania. 
Conservation efforts are consequently hampered by the lack of detailed data on the distribution, behaviour and ecology of blue whales, particularly during the winter and for the juvenile phase of life. 
An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotopes\cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
Stable isotope data therefore provide information on diet and location, but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values. 
Here we reconstruct movements of a historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information that are often under-utilised \cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.

Mysticete (baleen) whales are characterized by the development of baleen plates, keratinous structures in the upper jaw used to filter food items from seawater. 
Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert\cite{best1996stable}. 
Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although baleen is worn away at the tips over time, so rarely records an individual's entire lifespan. 
Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales\cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the north Atlantic population. 
Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope''. 
Hope is estimated to be a young adult c.15 years at death, suggesting that her baleen may yield information about juvenile behaviours in north Atlantic blue whales around the peak of industrial whaling.

Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on around three weeks prior to that time (this time is required for the carbon and nitrogen to be assimilated and converted into tissues). %REF???
% add line about it being ok to use phytoplankton?

We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth, resulting in 97 discrete samples of baleen.
We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin.
These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e. relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope\cite{west2006stable}.
Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; both isotopes provide information on the geographical locations of an individual, but nitrogen isotopes are more strongly fractionated during protein metabolism and thus contain additional trophic and physiological information\cite{west2006stable}.

Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplementary Methods). 
Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$, we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over seven full years of the whale's life (early 1884 - spring 1891). 
The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Fig. \ref{fig1}; Figs. S1 and S2). 
In behavioural phase two (summer 1886 to spring 1891) $\delta^{15}$N values are relatively elevated and $\delta^{13}$C values are relatively depleted with coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values. 
In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively positive $\delta^{13}$C values in the second half of 1891. 
The final three months of the record show a progressive fall in $\delta^{13}$C values (Fig. \ref{fig1}; Figs. S1 and S2).

% figure 1
\begin{figure}
  \centering
  \includegraphics[width = \linewidth]{figures/Figure-1-raw-dC-dN-data.png}
  \caption{Variation in stable isotope values in the NHM blue whale, expressed as $\delta^{13}$C (black circles, left y-axis) and $\delta^{15}$N (grey circles, right y-axis). Samples were taken longitudinally through the baleen plate (n = XX samples from a single baleen plate for both isotopes). There is strong annual periodicity and cross-correlation (Figs. S1 and S2) in both isotopes. The approximate relationship to years assuming a growth rate of 13.5cm y$^{-1}$ is shown on the upper x-axis, and year boundaries are indicated by vertical dotted grey lines.}
  \label{fig1}
\end{figure}
